Plume-SPH 1.0: A 3D dust-gas volcano plume model based on smoothed particle hydrodynamics

### 2016-10-04 Manuscripts_GMD.tex
slight modification

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh-based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh-based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Affect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementation of SPH. Core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plum models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our concern in current work is the risks that volcanic ash posed on aircrafts and communities. During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days. Those VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide sources terms for VATDs. This paper reports a new 3D (three dimensional) volcano plume model exploiting advantage of mesh-free methods in multiple phase free boundary flow modeling.\\

@@

The basic physics behind of volcanic plume has been understood for more than half a century \citep{morton1956turbulent, settle1978volcanic, wilson1978control}. Several volcano plume models have been developed in the past few decades. Specific aspects of a plume have been studied by 1D (one dimensional) plume models which describe the steady state solution of a plume under idealized boundary conditions.
The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details through 1D numerical simulations. Recently developed 1D models (eg. \citep{mastin2007user, woodhouse2013interaction}) tend to accounting for more physics effects in one model. FPLUME-1.0 \citep{folch2016fplume} accounts for wind affect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, In these 1D models, the entrainment of air is evaluated based two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models requires little computational sources and can run on standard personal computers or on web sites in very short time. As a result, 1D software (such as \citep{267, 1194, 3541}) for volcano plume development, combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skills in associated hazards forecasts with 1D models requires 2D (two dimensional) and 3D models.\\

@@

While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PADC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. An further extension was made to include particle aggregation in 2006 \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still pretty coarse compared with SK-3D and PDAC.
Besides strengthening their strength (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC is trying to include effect of microphysics into the model while ATHAM wants to extend its ability to modeling pyroclastic flow. Both are using finer and finer resolution. Recently, A first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, is derived based on three dimensional N phases Eulerian transportation equation. ASHEE is valid for low concentration and low Strokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adopting unstructured finite volume method.\\
To summarize, each model has its own focus based on problem of interest. Different simplification was made based on their scope. Improvements have been making since their birth. Accuracy (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) are always a couple of conflict considerations in 3D plume simulations.\\

@@

All of these existing 3D models are based on mesh-based methods (Eulerian methods). And there is no 3D plume model based on mesh free method (Lagrange methods) yet. Actually, Lagrangian methods have several features that are suitable for volcano plume simulation. Among such Lagrangian methods, Smoothed Particle Hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, even though relatively newer compared to well designed mesh based methods such as Finite Difference, has shown good agreement with experiments by many applications in fluid dynamic area. And it currently is, by far the most widely used scheme. Specifically, the reason we choose SPH as the numerical method for volcanic plume simulation are as following:

@@

\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiple phase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraph, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.

@@

The physics model of the plume is first presented in section \ref{sec:physics-model} and leads to a complete mathematical description of the volcano plume (governing equations and boundary conditions). In section \ref{sec:SPH_method} we will briefly introduce the numerical tool, SPH method. Both the fundamental discretization formulism and techniques that used to handle specific issues involved in plume modeling will be discussed. Validation and verification with numerical tests is presented in section \ref{sec:verification-validation}. In section \ref{sec:conclusion}, a discussion on future work is given following a brief summary.

@@

During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising kilometers to tens of kilometers until its heat is exhausted that buoyancy is lost. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. Simutaneously, relative coarse particle might separate with main stream of the plume falling down onto the ground and may be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor will condense to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor will further heat up entrained air and make the bulk density further dilute. Even entrained vapor might also experience similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation. 
This is essentially a multiple phases turbulent mixing process coupled with heat transfer and many other microphysics. As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and physics heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases will not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term will not show up with this assumption). Because of above assumptions, all other micro-physics process (like phase change of  \texorpdfstring{H\textsubscript{2}O}, aggregation, decomposition, absorption of gas on the surface of solid, solution of gas into the liquid) and chemical process will not be considered in this model. These microphysics factors will come to play critical roles under particular eruption conditions. Capturing of these processes need a more comprehensive model. One critical element in plume development, the affect of wind, is also not considered in this model yet. Introducing of wind effect in our model will need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind affect plays a significant role in its development, usually refereed as weak plume. Our model is also lack ability in modeling plumes with large particles or in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano at April 4th \citep{sigurdsson19841982, folch2016fplume}. At present stage, we are more interested in more basic (and more critical) aspects of volcano plume and therefor devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods.\\

@@

$E = e + K $ is total energy which is summation of kinetic energy $K$ and internal energy $e$.

@@

\section{SPH Method} \label{sec:SPH_method}
SPH is a mesh free Lagragian method. In SPH, the domain is discretized as a set of particles or discretization points and the position of each particle (discretization point) is updated at every time step. Approximation of all field variables (velocity, density and pressure, ect.) is obtained by interpolation based on discretization points. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of PDEs (partial differential equations) need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables at any point, the integral equations are evaluated as sums over neighbor particles. Thus, physical properties (density, velocity, internal energy, ect.) associated to the particle are updated based on its neighbors. The kernel function has a compact support determined by smoothing length of each particle. There are several review papers \citep{monaghan1992smoothed, monaghan2005smoothed, rosswog2009astrophysical, price2012smoothed, monaghan2012smoothed}, giving a pretty comprehensive view over SPH, we will only refer here to the representation of the constitutive equations in SPH and focus on numerical techniques for plume modeling.

@@

There are several routines for discretizing governing equations (partial differential equations or ordinary differential equations) with SPH. There is slight difference between discretized formulism obtained by different routines. We present here one of them following JJ Monaghan \citep{monaghan1992smoothed, monaghan2005smoothed}. The start point of approximating a function with SPH is one translation property of Dirac function.

@@

Recall that $d\textbf{x}\prime = \dfrac{dm \prime}{\rho \prime}$, the integration equation, Eq. (\ref{eq:SPH-fundamental-principle}), can be approximated by summation and lead to an approximation of the function at a particle (discretization point) a:

@@

Where the summation is over all the particles b within the region of compact support of the weighting function. 

@@

For vector quantities the expressions are similar, simply replacing $A$ with $\textbf{A}$ in Eq. (\ref{eq:SPH-approximation-sum}), giving

@@

%$just Italic for math symbol A$ \\
%$\textbf{bold Italic for vector A}$\\
%$\textbf{\text{bold norm for matrix A}}$ \\
%\textit{ Italic \textbf{bold Italic, A for vector}}
%\textbf{For matrix, Just bold A}

@@

Where, $\textbf{v}_{a b} = \textbf{v}_a - \textbf{v}_b$. $\Pi$ is artificial viscosity term, will be discussed in section \ref{sec:artificial-viscosity}.

@@

Adding new physics and new phases into the model is trivial in terms of discretization. For example, add of new source (or sink) into Eq.  (\ref{eq:ns-sph-d}), add a drag force into Eq. (\ref{eq:ns-sph-v})  and a heat exchange term into Eq. (\ref{eq:ns-sph-e}) will lead to the new discretization form:

@@

Where the source term $\dot{\rho}$ can be a "sink" of erupted vapor due to phase change which is considered in ATHAM \citep{oberhuber1998volcanic}. $D$ is drag force coefficient. $\kappa$ is heat conduction coefficient. Other physics can be added easily in a similar way. Adding of these new terms will lead to modification at several lines in the source code. The drag force term should show up only when dynamics dis-equilibrium between different phases is considered. In that case, each phase will need one set of governing equations of Navier-Strokes type. Adding of new phases into SPH code will only need add several new lines for the new phase besides adding interaction terms.

@@

In classic SPH, shock waves are handled by introducing artificial viscosity to smear out discontinuities. $\Pi$, in discretized momentum equation, Eq. (\ref{eq:ns-sph-v}), and energy equation, Eq. (\ref{eq:ns-sph-e}), represents artificial viscosity term which is essentially a second order derivative term. As in the case of first derivatives, second derivatives can be estimated by differentiating a SPH interpolation twice. However, such formulism has two disadvantages: First, it is very sensitive to particle disorder. Second, the second derivative of the kernel can change sign and will lead to unphysical representation (viscous dissipation cause decrease of the entropy). 

@@

\subsection{multiple phase SPH}
Numerical simulation of multiphase flows is usually difficult due to the existence of interface between two fluids. And the interface is usually moving or deformable. This present severe challenges to conventional Eulerian grid-based numerical methods. As we discussed in the introduction section, special algorithms to treat and track the interface between different phases is required for mesh-based method to deal with multiple phase problems. What's worse, the engulfment process during plume development makes geometry of interface very complicated. SPH is able to handle multiple phase problems without additional numerical effort for dealing with interface, which makes it suitable for volcano plume simulation. Take the governing equation adopted in our model as an example. Different phases of fluid (air and erupted material) will be represented by two different sets of SPH particles (or discretization points). Based on assumption we made in section \ref{sec:physics-model}, only density need to be updated respectively for each phase. The updating of density is exactly the same as Eq. (\ref{eq:ns-sph-d}) in spirit.  Particles of phase 1 will not be counted while evaluate density of phase 2 and vice verse. Updating of density is based on the following discretized equations.

@@

<\rho_{\alpha}^a>=\frac{\sum m_b w_{\alpha b} (h_{\alpha})}{\sum \frac{m_b}{\rho_b} w_{\alpha b} (h_{\alpha}) +\sum \frac{m_j}{\rho_j} w_{\alpha j} (h_\alpha)} \label{eq:gov-sph-d1}

@@

<\rho_\alpha^{sg}>=\frac{\sum_j m_j w_{\alpha j} (h_\alpha)}{\sum \frac{m_b}{\rho_b} w_{\alpha b} (h_{\alpha}) +\sum \frac{m_j}{\rho_j} w_{\alpha j} (h_\alpha)} \label{eq:gov-sph-d2}

@@

$\rho=\rho^a + \rho^{sg}$ is density of mixture of phase 1 and phase 2. In the areas far away from interface, updating of density is exactly the same as that for single phase flow. For example, in the right side and left side (or blue areas) in Fig. \ref{fig:SPH-multiple-density}, where there are only air particles, Eq.  (\ref{eq:gov-sph-d2}) will lead to zero and total density will be:

@@

That is to say, the same density updating equation can be applied for both phases and no additional numerical treatment needed to locate where is the interface.

@@

Another way is to use extremely fine resolution (The direct numerical simulation method) at the expense of much higher computational cost that results from both increase in number of particles and decrease in time steps constrained by CFL condition. Here we choose the SPH-SPS method to capture  momentum and energy exchange caused by smaller scale turbulence.\\

>>>>

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Effect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementation of SPH. Core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plum models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our concern in current work is the risks that volcanic ash posed on aircrafts and communities. During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days. Those VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide sources terms for VATDs. This paper reports a new 3D (three dimensional) volcano plume model exploiting advantage of mesh-free methods in multiphase free boundary flow modeling.\\

@@

The basic physics of volcanic plume has been understood for more than half a century \citep{morton1956turbulent, settle1978volcanic, wilson1978control}. Several volcano plume models have been developed in the past few decades. Specific aspects of a plume have been studied by 1D (one dimensional) plume models which describe the steady state solution of a plume under idealized boundary conditions.
The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details through 1D numerical simulations. Recently developed 1D models (eg. \citep{mastin2007user, woodhouse2013interaction}) tend to account for more physics effects in one model. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, In these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models requires little computational sources and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcano plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skills in associated hazards forecasts with 1D models requires 2D (two dimensional) and 3D models.\\

@@

While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PADC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation in 2006 \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still pretty coarse compared with SK-3D and PDAC.
Besides strengthening their strength (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC is trying to include effect of microphysics into the model while ATHAM wants to extend its ability to modeling pyroclastic flow. Both are using finer and finer resolution. Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, is derived based on three dimensional N phases Eulerian transportation equation. ASHEE is valid for low concentration and low Strokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adopting unstructured Finite Volume method.\\
To summarize, each 3D model has its own focus based on problem of interest. Different simplification was made based on their scope. Improvements have been making since their birth. Accuracy (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) are always a couple of conflict considerations in 3D plume simulations.\\

@@

All of these existing 3D models are based on mesh based methods (Eulerian methods). And there is no 3D plume model based on mesh free method (Lagrange methods) yet. Actually, Lagrangian methods have several features that are suitable for volcano plume simulation. Among such Lagrangian methods, Smoothed Particle Hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, even though relatively newer compared to well designed mesh based methods such as Finite Difference, has shown good agreement with experiments by many applications in fluid dynamic area. And it currently is, by far the most widely used scheme. Specifically, the reason we choose SPH as the numerical method for volcanic plume simulation are as following:

@@

\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraph, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.

@@

The physics model of the plume is first presented in section \ref{sec:physics-model} and leads to a complete mathematical description of the volcano plume (governing equations and boundary conditions). In section \ref{sec:SPH_method} we will briefly introduce the numerical tool, SPH method. Both the fundamental discretization formulism and techniques that used to handle specific issues involved in plume modeling will be discussed. Verification and validation with numerical tests is presented in section \ref{sec:verification-validation}. In section \ref{sec:conclusion}, a discussion on future work is given following a brief summary.

@@

During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising kilometers to tens of kilometers until its heat is exhausted that buoyancy is lost. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. Simutaneously, relative coarse particle might separate with main stream of the plume falling down onto the ground and may be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor will condense to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor will further heat up entrained air and make the bulk density further dilute. Even entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation. 
All in all, the process of plume development is  essentially a multiphase turbulent mixing process coupled with heat transfer and many other microphysics and chemical reaction. As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and physics heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases will not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term will not show up with this assumption). Because of above assumptions, all other micro-physics process (like phase change of  \texorpdfstring{H\textsubscript{2}O}, aggregation, decomposition, absorption of gas on the surface of solid, solution of gas into the liquid) and chemical process will not be considered in this model. These ignored microphysics factors will come to play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model will need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refereed as weak plume. Our model is also lack ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano at April 4th \citep{sigurdsson19841982, folch2016fplume}. At present stage, we are more interested in more basic (and more critical) aspects of volcano plume and therefor devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods.\\

@@

$E = e + K $ is total energy which is a summation of kinetic energy $K$ and internal energy $e$.

@@

\section{SPH method} \label{sec:SPH_method}
SPH is a mesh free Lagragian method. In SPH, the domain is discretized as a set of particles or discretization points and the position of each particle (discretization point) is updated at every time step. Approximation of all field variables (velocity, density and pressure, ect.) is obtained by interpolation based on discretization points. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of PDEs (partial differential equations)  or ODEs (ordinary differential equations) need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables at any point, the integral equations are evaluated as sums over neighbor particles. Thus, field variables associated to the particle are updated based on its neighbors. The kernel function has a compact support determined by smoothing length of each particle. There are several review papers \citep{monaghan1992smoothed, monaghan2005smoothed, rosswog2009astrophysical, price2012smoothed, monaghan2012smoothed}, giving a pretty comprehensive view over SPH, we will only refer here to the representation of the constitutive equations in SPH and put more focus on specific numerical techniques for plume modeling.

@@

There are several routines for discretizing governing equations (PDEs or ODEs) with SPH. There is slight difference between discretized formulism obtained by different routines. We present here one of them following JJ Monaghan \citep{monaghan1992smoothed, monaghan2005smoothed}. The start point of approximating a function with SPH is one translation property of Dirac function.

@@

Recall that $d\textbf{x}\prime = \dfrac{dm \prime}{\rho \prime}$, the integration equation, Eq. (\ref{eq:SPH-fundamental-principle}), can be approximated by summation and lead to an approximation of the function at particle (discretization point) a:

@@

Where the summation is over all the particles within the region of compact support of the weighting function. 

@@

For vector quantities the expressions are similar, simply replacing $A$ with $\textbf{A}$ in Eq. (\ref{eq:SPH-approximation-sum}) and and Eq. (\ref{eq:SPH-scalar-function-gradient}), giving

@@

@@

Where, $\textbf{v}_{a b} = \textbf{v}_a - \textbf{v}_b$. $\Pi$ is artificial viscosity term, which will be discussed in section \ref{sec:artificial-viscosity}.

@@

Adding new physics and new phases into the model is trivial in terms of discretization. For example, adding of new source (or sink) into Eq.  (\ref{eq:ns-sph-d}), adding a drag force into Eq. (\ref{eq:ns-sph-v})  anding a heat exchange term into Eq. (\ref{eq:ns-sph-e}) will lead to the new discretization form:

@@

Where the source term $\dot{\rho}$ can be a "sink" of erupted vapor due to phase change which is considered in ATHAM \citep{oberhuber1998volcanic}. $D$ is drag force coefficient. $\kappa$ is heat conduction coefficient. Other physics can be added easily in a similar way. Adding of these new terms will lead to modification at several lines in the source code. The drag force term should show up only when dynamics dis-equilibrium between different phases is considered. In that case, each phase will need one set of governing equations of Navier-Strokes type. Adding of new phases into SPH code only needs to add several new lines for the new phase besides adding interaction terms.

@@

In classical SPH, shock waves are handled by introducing artificial viscosity to smear out discontinuities. $\Pi$, in discretized momentum equation, Eq. (\ref{eq:ns-sph-v}), and energy equation, Eq. (\ref{eq:ns-sph-e}), represents artificial viscosity term which is essentially a second order derivative term. As in the case of first derivatives, second derivatives can be estimated by differentiating a SPH interpolation twice. However, such formulism has two disadvantages: First, it is very sensitive to particle disorder. Second, the second derivative of the kernel can change sign and will lead to unphysical representation (viscous dissipation cause decrease of the entropy). 

@@

\subsection{Multiphase SPH}
Numerical simulation of multiphase flows is usually difficult due to the existence of interface between two fluids. And the interface is usually moving or deformable. This present severe challenges to conventional Eulerian grid-based numerical methods. As we discussed in the introduction section, special algorithms to treat and track the interface between different phases is required for mesh based method to deal with multiphase problems. What's worse, the engulfment process during plume development makes geometry of interface very complicated. SPH is able to handle multiphase problems without additional numerical effort for dealing with interface, which makes it suitable for volcano plume simulation. Take the governing equation adopted in our model as an example. Different phases of fluid (air and erupted material) will be represented by two different sets of SPH particles (or discretization points). Based on assumption we made in section \ref{sec:physics-model}, only density need to be updated respectively for each phase. The updating of density is exactly the same as Eq. (\ref{eq:ns-sph-d}) in spirit.  Particles of phase 1 will not be counted while evaluate density of phase 2 and vice verse. Updating of density is based on the following discretized equations.

@@

<\rho_{\alpha}^a>=\frac{\sum m_b w_{\alpha b} (h_{\alpha})}{\sum \frac{m_b}{\rho_b} w_{\alpha b} (h_{\alpha}) +\sum \frac{m_j}{\rho_j} w_{\alpha j} (h_{\alpha})} \label{eq:gov-sph-d1}

@@

<\rho_\alpha^{sg}>=\frac{\sum_j m_j w_{\alpha j} (h_\alpha)}{\sum \frac{m_b}{\rho_b} w_{\alpha b} (h_{\alpha}) +\sum \frac{m_j}{\rho_j} w_{\alpha j} (h_{\alpha})} \label{eq:gov-sph-d2}

@@

$\rho=\rho^a + \rho^{sg}$ is density of mixture of air and erupted material. By definition, the mass fraction is updated according to Eq. (\ref{eq:gov-sph-xi}).
<\xi_{\alpha}> = \dfrac{\rho^{sg}_{\alpha}}{\rho_{\alpha}}
\label{eq:gov-sph-xi}
In the areas far away from interface, updating of density is exactly the same as that for single phase flow. For example, in the right side and left side (or blue areas) in Fig. \ref{fig:SPH-multiple-density}, where there are only air particles, Eq.  (\ref{eq:gov-sph-d2}) will lead to zero and total density will be:

@@

That is to say, the same density updating equation can be applied for both phases and no additional numerical treatment needed to locate where the interface is.

@@

Another way is to use extremely fine resolution (The direct numerical simulation method) at the expense of much higher computational cost that results from both increase in number of particles and decrease in time steps constrained by CFL condition. Here we choose the SPH-SPS method.\\
